% main.tex — P03: Zigzag Persistence, Survey Design, and Labour Market Fragmentation
% Sociological Methods & Research submission / arXiv preprint
%
% Build: make          (runs pdflatex + bibtex twice)
%        latexmk -pdf main.tex  (continuous)

\documentclass[12pt, a4paper]{article}

% --- Encoding & fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathpazo}           % Palatino for body text
\usepackage{inconsolata}        % monospace

% --- Page geometry ---
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\doublespacing                  % journal requirement

% --- Mathematics ---
\usepackage{amsmath, amssymb, amsthm}

% --- Tables ---
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{longtable}

% --- Graphics ---
\usepackage{graphicx}
\graphicspath{{figures/}{../figures/}}
\usepackage[font=small, labelfont=bf]{caption}
\usepackage{subcaption}

% --- Hyperlinks (load last among basics) ---
\usepackage[colorlinks=true, linkcolor=blue!60!black, citecolor=blue!60!black, urlcolor=blue!60!black]{hyperref}

% --- Bibliography ---
\usepackage[round, authoryear]{natbib}
\bibliographystyle{apalike}

% --- Misc ---
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{microtype}          % better line-breaking

% --- Custom commands ---
\newcommand{\HH}[1]{H_{#1}}    % homology dimension
\newcommand{\PH}{\mathrm{PH}}
\newcommand{\VR}{\mathrm{VR}}
\newcommand{\Wass}{W_2}         % 2-Wasserstein distance
\newcommand{\pval}{\textit{p}}  % p-value formatting
\newcommand{\placeholder}[1]{\textbf{[#1]}}  % mark placeholders
\newcommand{\TP}{\mathrm{TP}}   % total persistence
\newcommand{\AUC}{\mathrm{AUC}} % area under Betti curve
\newcommand{\PE}{\mathrm{PE}}   % persistence entropy

% --- Title ---
\title{Topological Business Cycles?\\
  Zigzag Persistence, Survey Design, and Labour Market\\
  Fragmentation in the UK, 1991--2022}
\author{%
  Stephen Dorman\\
  \textit{The Open University, UK}
}
\date{March 2026}

% ============================================================
\begin{document}
\maketitle

\begin{abstract}
We apply zigzag persistent homology to 32 annual snapshots of UK
employment-income trajectories from the British Household Panel Study and
Understanding Society ($N = 35{,}789$ individuals, 1991--2022).
Zigzag persistence tracks topological features across a temporal sequence of
spaces; each year's point cloud is a sub-sample of a common PCA embedding of
nine-state career $n$-gram vectors, with frozen loadings ensuring cross-year
comparability.

The na\"ive zigzag barcode tells an appealing story: a connected component born
at the 2008 Global Financial Crisis persists through 2022, while the 1993
recession produced only a transient 5.5-year fragment.
Pairwise Wasserstein distances between the 32 annual persistence diagrams
reinforce this picture, with a sharp two-block structure separating BHPS-era
(1991--2008) from USoc-era (2009--2022) topology.

We then show that this narrative is mostly wrong.
Era-specific zigzag runs, spanning-individual analysis, sub-sampled Betti curves
at matched sample sizes, and pool-draw null models converge on the same
conclusion: the dominant post-2008 topological shift is driven by Understanding
Society's expanded sampling frame, not by the GFC.
Spanning individuals present in both surveys maintain stable BHPS-like topology;
the 21,000 USoc-only newcomers, who disproportionately occupy previously
unsampled regions of the embedding, account for virtually all of the cross-era
topological difference.
The block structure in the Wasserstein matrix is highly significant under both
label-permutation (all six test statistics at $p < 0.001$) and pool-draw null
models ($p < 0.001$ for block ratio), confirming that the era separation is
real---but its source is survey design, not economic shock.

Within the BHPS era alone, where survey design is constant, total $\HH{0}$
persistence increases at $+0.31$/year ($p < 0.0001$), and this trend correlates
strongly with labour market composition variables---employment rate ($r = +0.82$),
part-time share ($r = +0.80$), unemployment ($r = -0.81$)---but not with GDP
growth.
First-differenced correlations yield no significant associations after
multiple-testing correction, indicating that topology tracks slow structural
change in employment composition, not business-cycle dynamics.

The paper develops a diagnostic toolkit---spanning-individual decomposition,
era-specific robustness, pool-draw null models---for adjudicating between
substantive and artefactual topological change in linked panel data.
It documents a genuine secular increase in the topological complexity of UK
career trajectories during 1991--2008, consistent with labour market
flexibilisation, already underway before the financial crisis.

\medskip
\noindent\textbf{Keywords:} zigzag persistence, topological data analysis,
survey design artefacts, labour market trajectories, Wasserstein distance,
time-varying topology

\end{abstract}

\newpage
\tableofcontents
\newpage

% ============================================================
% Body content converted from Markdown draft v2-2026-03.md
\input{body}

% ============================================================
\bibliography{references}

\end{document}
